---
title: "Wildfire Risk Demo (Quarto)"
subtitle: "NW Iowa + SE South Dakota"
---

# Model Map {orientation="columns" scrolling="false"}

## Column {width=68%}

```{r setup, include=FALSE}
library(leaflet)
library(jsonlite)
library(sf)
library(dplyr)

meta_path <- "analysis_output/model_table_meta.json"
metrics_path <- "analysis_output/model_metrics.json"

meta <- if (file.exists(meta_path)) fromJSON(meta_path) else NULL
metrics <- if (file.exists(metrics_path)) fromJSON(metrics_path) else NULL

fmt_int <- function(x) {
  if (is.null(x) || length(x) == 0 || is.na(x)) return("n/a")
  format(as.integer(round(as.numeric(x))), big.mark = ",", scientific = FALSE)
}
```

```{r map, echo=FALSE, warning=FALSE, message=FALSE}
read_chunk_layer <- function(index_path, risk_col_candidates) {
  idx <- fromJSON(index_path, simplifyDataFrame = TRUE)
  rel <- vapply(idx$features$properties$path, identity, character(1))
  files <- file.path("docs", sub("^\\./", "", rel))
  files <- files[file.exists(files)]
  if (length(files) == 0) {
    return(NULL)
  }
  out <- lapply(files, function(p) st_read(p, quiet = TRUE)) |>
    bind_rows()
  if (is.na(st_crs(out))) {
    st_crs(out) <- 4326
  } else {
    out <- st_transform(out, 4326)
  }
  found_col <- NULL
  for (nm in risk_col_candidates) {
    if (nm %in% names(out)) {
      found_col <- nm
      break
    }
  }
  if (is.null(found_col)) {
    out$risk_display <- 0
  } else {
    out$risk_display <- as.numeric(out[[found_col]])
  }
  out
}

county_path <- "data/demo_counties.shp"
if (!file.exists(county_path)) {
  stop("County shapefile not found at data/demo_counties.shp")
}
counties_sf <- st_read(county_path, quiet = TRUE)
if (is.na(st_crs(counties_sf))) {
  st_crs(counties_sf) <- 4326
} else {
  counties_sf <- st_transform(counties_sf, 4326)
}
counties_sf <- st_make_valid(counties_sf)
county_mask <- st_union(counties_sf)
county_bbox <- st_bbox(counties_sf)

clip_to_counties <- function(x) {
  if (is.null(x) || nrow(x) == 0) return(x)
  x <- st_make_valid(x)
  x <- suppressWarnings(st_intersection(x, county_mask))
  if (nrow(x) == 0) return(NULL)
  x
}

forecast_sf <- read_chunk_layer(
  "docs/data/forecast_chunks/index.geojson",
  c("risk_display", "risk_weekly_max", "risk_index")
)
forecast_sf <- clip_to_counties(forecast_sf)

if (is.null(forecast_sf)) {
  stop("No forecast chunk layer found for Quarto map.")
}

all_vals <- c(
  if (!is.null(forecast_sf)) forecast_sf$risk_display else numeric(0)
)
if (length(all_vals) == 0) all_vals <- c(0, 100)

pal <- colorBin(
  palette = c("#2b83ba", "#abdda4", "#fdae61", "#d7191c"),
  bins = c(0, 25, 50, 75, 100),
  domain = all_vals,
  right = TRUE
)

m <- leaflet(options = leafletOptions(zoomControl = TRUE, minZoom = 6, maxZoom = 14)) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addWMSTiles(
    baseUrl = "https://gibs.earthdata.nasa.gov/wms/epsg3857/best/wms.cgi",
    layers = "MODIS_Terra_NDVI_8Day",
    options = WMSTileOptions(format = "image/png", transparent = TRUE, opacity = 0.45),
    group = "NDVI (NASA GIBS)"
  ) |>
  addPolylines(
    data = st_boundary(counties_sf),
    group = "County outlines",
    color = "#2f2f2f",
    weight = 2.0,
    opacity = 0.85
  ) |>
  fitBounds(
    lng1 = county_bbox[["xmin"]], lat1 = county_bbox[["ymin"]],
    lng2 = county_bbox[["xmax"]], lat2 = county_bbox[["ymax"]]
  )

if (!is.null(forecast_sf)) {
  m <- m |>
    addPolygons(
      data = forecast_sf,
      group = "Next-week forecast grid",
      color = "#9a9a9a",
      weight = 0.10,
      fillColor = ~pal(risk_display),
      fillOpacity = 0.42,
      popup = ~paste0(
        "<b>Cell:</b> ", ifelse(is.na(cell_id), "n/a", cell_id), "<br>",
        "<b>Forecast:</b> ", ifelse(is.na(forecast_start), "n/a", forecast_start),
        " to ", ifelse(is.na(forecast_end), "n/a", forecast_end), "<br>",
        "<b>Risk (display):</b> ", ifelse(is.na(risk_display), "n/a", round(risk_display, 1)), "<br>",
        "<b>Risk (weekly max):</b> ", ifelse(is.na(risk_weekly_max), "n/a", risk_weekly_max)
      )
    )
}

m |>
  addLayersControl(
    overlayGroups = c(
      "Next-week forecast grid",
      "County outlines",
      "NDVI (NASA GIBS)"
    ),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  addLegend(
    "bottomright",
    pal = pal,
    values = all_vals,
    title = "Risk legend",
    opacity = 0.9,
    labFormat = labelFormat(transform = function(x) round(x))
  ) |>
  fitBounds(
    lng1 = county_bbox[["xmin"]], lat1 = county_bbox[["ymin"]],
    lng2 = county_bbox[["xmax"]], lat2 = county_bbox[["ymax"]]
  )
```

## Column {width=32%}

### Plain-English Process

This tool maps where fire-conducive weather conditions are strongest in your counties.

1. Historical fire records are ingested for the area and date range.
2. A training table is built with positive fire points and matched non-fire controls.
3. Weather stress features are computed (hotter, drier, windier, rain deficit = higher risk).
4. The model is trained and validated with a time holdout.
5. Forecast weather is converted into a 1 km risk surface and clipped to county boundaries.

### Training Data Summary

```{r training-summary, echo=FALSE, results='asis'}
if (is.null(meta)) {
  cat("- Training metadata file not found (`analysis_output/model_table_meta.json`).\n")
} else {
  yr <- meta$year_range
  yr_txt <- if (!is.null(yr) && length(yr) == 2) paste0(yr[[1]], " to ", yr[[2]]) else "n/a"
  cat("- Training period:", yr_txt, "\n")
  cat("- Positives after dedupe:", fmt_int(meta$positives_after_firms_dedupe), "\n")
  cat("- Controls generated:", fmt_int(meta$control_rows), "\n")
  cat("- Total rows in model table:", fmt_int(meta$total_rows), "\n")
  cat("- Season filter:", ifelse(is.null(meta$season_filter), "n/a", meta$season_filter), "\n")
}
if (!is.null(metrics)) {
  cat("- Train rows used in latest fit:", fmt_int(metrics$train_rows), "\n")
  cat("- Test rows used in latest fit:", fmt_int(metrics$test_rows), "\n")
}
```

### Layer Meaning

- `Next-week forecast grid`: forward-looking projected risk for the selected forecast window.
- `County outlines`: your `demo_counties.shp` boundaries.
- `NDVI`: NASA GIBS vegetation-condition backdrop.

### Methods and Weighting

Weather score weights:

- temperature stress = `0.32`
- humidity stress = `0.23`
- wind stress = `0.25`
- rain-deficit stress = `0.20`

Display uses a regional linear stretch between p05 and p95 (not rank remapping).

### Input Data Sources

- [Local area boundary (your shapefile): `data/demo_counties.shp`](https://desktop.arcgis.com/en/arcmap/latest/manage-data/shapefiles/what-is-a-shapefile.htm)
- [Open-Meteo Archive and Forecast APIs (weather inputs)](https://open-meteo.com/en/docs)
- [NASA GIBS MODIS Terra NDVI 8-day layer (vegetation backdrop)](https://nasa-gibs.github.io/gibs-api-docs/)
- [NASA FIRMS active fire detections (historical fire input)](https://firms.modaps.eosdis.nasa.gov/)
- [NIFC historic wildfires source](https://data-nifc.opendata.arcgis.com/)
- [NWS reference products for operational cross-checking](https://www.weather.gov/)

# About {orientation="columns" scrolling="false"}

This Quarto view is presentation-only and does not rerun model training or generation scripts.
